---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<style> 
  /* https://stackoverflow.com/questions/31753897/2-column-section-in-r-markdown */
  .col2 {
    columns: 2 200px;         /* number of columns and width in pixels*/
    -webkit-columns: 2 200px; /* chrome, safari */
    -moz-columns: 2 200px;    /* firefox */
  }
  .col3 {
    columns: 3 100px;
    -webkit-columns: 3 100px;
    -moz-columns: 3 100px;
  }
  .column-left{
    float: left;
    width: 30%;
    text-align: left;
  }
  .column-right{
    float: right;
    width: 70%;
    text-align: right;
  }
  /* https://gist.github.com/jennybc/e9e9aba6ba18c72cec26 */
  .clearer {clear: both}
  
  .center{
    text-align: center
  }
</style>

# diftrans

<!-- badges: start -->
![status](https://img.shields.io/badge/status-under%20construction-yellow)
<!-- badges: end -->

The goal of the `diftrans` package is to compute the differences-in-transports estimator introduced in Daljord, Pouliot, Hu, and Xiao (2020). Currently, the package is in development.

## Installation

You can install the `diftrans` from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("omkarakatta/diftrans")
```
## Example

The workhorse function in the `diftrans` package is `get_results`, which serves two purposes:

* compute transport cost between two univariate distributions, and
* compute the differences-in-transports estimator (see Daljord et al. (2020)).

We begin by computing the transport cost between two distributions, which we shall call
the pre-distribution and post-distribution. They should be represented as tibbles with
two columns:

* column 1 contains the full support of the distribution and 
* column 2 (labeled "count") contains the mass/counts associated with each value in the support.

### Setup

Below are the tibbles for the pre- and post-distributions as a well as corresponding
plot. 
Both tibbles contain the full support of our variable of interest, that is, their
`support` column is the same. 


```{r echo = F, message = F, warning = F}
library(dplyr)
library(magrittr)
library(ggplot2)
library(knitr)
devtools::load_all()
```

<div class="center"> <strong> Pre-Distribution </strong> </div>

<div class="column-left">
```{r echo = F}
set.seed(1)
support_min <- 1
support_max <- 10
support <- seq(support_min, support_max, 1)
pre_distribution <- data.frame(support = sample(support, 100, replace = T)) %>% 
  group_by(support) %>% 
  count() %>% 
  ungroup() %>% 
  rename(count = n)

shift <- 5
for (i in 1:shift){
  pre_distribution[support == support_max-i+1, "count"] <- 0
}

kable(pre_distribution)
```
</div>
<div class = "column-right">
```{r echo = F}
ggplot(pre_distribution, aes(x = support, y = count)) +
  geom_bar(stat = "identity") +
  theme_bmp(sizefont = NULL,
            axissizefont = NULL) +
  scale_x_continuous(breaks = support,
                     labels = support)
```
</div>
<div class = "clearer"> </div>

<div class="center"> <strong> Post-Distribution </strong> </div>

<div class="column-left">
```{r echo = F}
post_distribution <- data.frame(support = support,
                                count = lag(pre_distribution$count, shift)) %>% 
  replace_na(list(count = 0))

kable(post_distribution)
```
</div>
<div class = "column-right">
```{r echo = F}
ggplot(post_distribution, aes(x = support, y = count)) +
  geom_bar(stat = "identity") +
  theme_bmp(sizefont = NULL,
            axissizefont = NULL) +
  scale_x_continuous(breaks = support,
                     labels = support)
```
</div>
<div class = "clearer"> </div>

Observe that the post-distribution is simply the result of shifting the mass in 
the pre-distribution by `r shift` units. For instance, all the mass that was given 
given to `r pre_distribution[1, "support"]` in the pre-distribution is now given
to `r pre_distribution[1+shift, "support"]`.

### Compute Transport Cost

We can compute the transport cost as follows:

```{r echo = T}
tc <- get_results(pre_main = pre_distribution, post_main = post_distribution,
                  estimator = "tc", var = support,
                  bandwidth = 0)

tc
```

The transport cost is `r scales::percent(tc$main)`, which makes sense since all the mass in the pre-distribution moves to values of the support that did not have any mass in the post-distribution, i.e. each unit of mass was transported.
The `estimator` argument is specified to be `"tc"`, which stands for transport cost.
(Since we only have two distributions, estimator will take on the value of `"tc"` by default.)

Since the post-distribution is a `r shift`-unit shift in the pre-distribution, any bandwidth
at least `r shift` must result in a transport cost of 0. We verify this by computing
the transport cost for a sequence of bandwiths from 0 to 10:
```{r}
tc <- get_results(pre_main = pre_distribution, post_main = post_distribution,
                  estimator = "tc", var = support,
                  bandwidth = seq(0, 10))

tc
```

### Setup for Difference-in-Transports

To provide a bit more context to our synthetic example, suppose that the shift 
in the post-distribution of our variable of interest is considered to be the result
of some treatment. Perhaps there are unobserved trends in our variable that might
also explain this shift. To elicit how much of this shift is due to the treatment
and how much of this shift is due to unobserved trends, we can compare our distributions
above with distributions of the same variable measured in some control group that
did not receive the treatment but was also subjected to the same unobserved trends.

The novel differences-in-transports estimator described in Daljord et al. (2020)
quantifies the change in distribution due to some treatment relative to the change
in distribution of some control group.

```{r echo = F}
control_shift <- 2

pre_control <- data.frame(support = support,
                          count = pre_distribution$count)
post_control <- data.frame(support = support,
                           count = lag(pre_distribution$count, control_shift)) %>% 
  replace_na(list(count = 0))
```

First, we need to set up the distributions of our variable for our control group,
both before and after the treatment was administered to the treatment group.
For our purposes, we let the pre-distribution of our control group be the same
as that of our treated group. Then, let our post-distribution for our control group
be a `r control_shift`-unit shift of our pre-distribution. We therefore have the 
following tibbles that represent our control group distributions:

<div class="center"> <strong> Pre-Distribution </strong> </div>

<div class="column-left">
```{r echo = F}
kable(pre_control)
```
</div>
<div class="column-right">
```{r echo = F}
ggplot(pre_control, aes(x = support, y = count)) +
  geom_bar(stat = "identity") +
  theme_bmp(sizefont = NULL,
            axissizefont = NULL) +
  scale_x_continuous(breaks = support,
                     labels = support)
```
</div>
<div class = "clearer"> </div>

<div class="center"> <strong> Post-Distribution </strong> </div>

<div class="column-left">
```{r echo = F}
kable(post_control)
```
</div>
<div class="column-right">
```{r echo = F}
ggplot(post_control, aes(x = support, y = count)) +
  geom_bar(stat = "identity") +
  theme_bmp(sizefont = NULL,
            axissizefont = NULL) +
  scale_x_continuous(breaks = support,
                     labels = support)
```
</div>
<div class = "clearer"> </div>

### Compute Differences-in-Transports Estimator

There is more nuance to computing the differences-in-transports estimator than 
is presented in this expository document. See Daljord et al. (2020) for a more 
complete treatment on how to compute the differences-in-transports estimator.

Above, we used `estimator = "tc"` and specified our pre- and post-distributions
for our treated group. 
Now, we use `estimator = "dit"` (which stands for differences-in-treatments estimator)
while also specifying our pre- and post-distributions for our control group.

```{r}
dit <- get_results(pre_main = pre_distribution, post_main = post_distribution,
                   pre_control = pre_control, post_control = post_control,
                   estimator = "dit", var = support,
                   bandwidth_seq = seq(0, 10, 1),
                   save_dit = TRUE)
dit$out
dit$dit
dit$optimal_bandwidth
```

Another difference between computing the transport cost and the differences-in-differences
estimator is that the differences-in-transports estimator is printed as a message. To
save this result, we use `save_dit = TRUE`.

Thus, the differences-in-transports estimator at an optimal bandwidth of `r dit$optimal_bandwidth`
is `r scales::percent(dit$dit)`. Often times, the data will be a sample from a population,
and sample variability is enough to cause differences between distributions. To account
for this discrepancy, we can increase the bandwidth to ignore transfers of mass between
nearby values of the support.

```{r}
d_a <- 3
```

While Daljord et al. (2020) offer a disciplined way to determine what the appropriate
bandwidths are, for simplicity, we will suppose that there will not be any sampling
variation with a bandwidth greater than `r d_a`. 

```{r}
dit <- get_results(pre_main = pre_distribution, post_main = post_distribution,
                   pre_control = pre_control, post_control = post_control,
                   estimator = "dit", var = support,
                   bandwidth_seq = seq(d_a, 10, 1), # smallest bandwidth will be 1
                   save_dit = TRUE)
dit$out
dit$dit
dit$optimal_bandwidth
```

Accounting for sampling variability, our differences-in-transports estimate is 
`r scales::percent(dit$dit)`. If we want to be more conservative, we can use
`conservative = TRUE`, which essentially uses twice the bandwidth for computing the 
treated group's transport costs relative to the bandwidth for computing the control 
group's transport costs.

```{r}
dit <- get_results(pre_main = pre_distribution, post_main = post_distribution,
                   pre_control = pre_control, post_control = post_control,
                   estimator = "dit", var = support,
                   bandwidth_seq = seq(d_a, 10, 1), # smallest bandwidth will be 1
                   save_dit = TRUE,
                   conservative = TRUE)
dit$out
dit$dit
dit$optimal_bandwidth
```

Finally, we have that the conservative differences-in-transports estimator is
`r scales::percent(dit$dit)` at an optimal bandwidth of `r dit$optimal_bandwidth`.


Note that `estimator = "dit"` is not necessary. Since we have two sets of pre- and
post-distributions, `get_results` is smart enough to return the differences-in-transports 
estimator by default. 

